---
title: "Final"
format: 
  html:
    self-contained: true
bibliography: references.bib
---

## Research Question

How does party fragmentation within a political system affect voter turnout in electoral democracies?

## Introduction

There are a number of studies in electoral research that have attempted to investigate the relationship between party fragmentation and measures of democratic strength. The direction of the effect, however, is widely disputed among political scientists. On the one hand, some scholars argue that greater fragmentation fosters pluralism, representation, and policy innovation by ensuring that a wider range of political interests are represented in the decision-making process. This, in turn, can enhance democratic legitimacy and participation. Other researchers contend that high levels of fragmentation can lead to governmental instability, inefficiency, and legislative gridlock, ultimately weakening democratic performance.

Multiple measures of democracy have been studied in the literature, including governmental stability, accountability, corruption, turnout, public goods provision, female representation and the representation of minorities [@valentim2024]. The empirical findings from these studies are often contradictory concerning the direction of the effect of party fragmentation on democratic quality, even when comparing the same indicators.

A number of studies have also focused on characterizing the specific relationship between party fragmentation and political participation, measured by voter turnout. A large group of scholars argues that party system fragmentation likely depresses turnout since citizens vote for a small part of the government which will likely form a coalition to rule. Additionally, each individual candidate has a lower likelihood of winning elections in a fragmented party system, which may also discourage voter turnout. Additionally, a higher number of parties increase the cost of voting by increasing information costs as one needs to research a higher number of candidates and policies to make a decision on who to vote for. It also creates a more complex system, which may put uneducated voters at a disadvantage. In empirical research, party fragmentation has also been found to be negatively correlated with voter turnout [@franklin1998; @geys2006; @henderson2010; @jackman1995; @kostadinova2007; @lehoucq2004; @radcliff2000].

However, there is also empirical evidence pointing to a positive relationship between party fragmentation and voter turnout [@banducci2009], though the authors find a stronger relationship when regressing on party polarization. The theoretical foundation behind this research is that a greater number of parties may increase voter turnout by providing more ideological options that align with citizens' preferences, thereby encouraging participation. Higher competition in the party system could also be associated with better policies that are more representative of the population, thus changing the voter’s cost-benefit calculation when deciding to vote. However, some scholars argue that an excessive number of parties can dilute ideological clarity and create confusion, potentially discouraging voter engagement.

All of the studies above that use country-level aggregate data focus on industrialized democracies. The justification for this is that those kinds of regimes are more consolidated democracies with very limited experience with authoritarianism and where the right to vote is generally guaranteed. Other studies have limited their research to a number of elections in a particular country, hoping to control for the endogeneity of political parties. Few studies have expanded the number of countries considered to all electoral democracies. @blais1998 assess the effect of many variables on turnout in 324 democratic national lower house elections held in 91 countries, between 1972 and 1995. They show that the number of parties has a statistically significant and negative effect on turnout. Though they conclude that the results did not differ dramatically from studies that consider a narrower sample of countries, the authors do not theorize about how party fragmentation might operate differently when influencing turnout in less established democracies. In fact, to my knowledge, no studies theorize about that distinction and use large-scale empirical analysis to test it. To fill this gap, I will propose that developing democracies have less institutionalized parties, which presents nuances to the way in which fragmentation can influence turnout.

Studies that focus on developing democracies are more case-study based and do not consider a large sample of countries. For instance, Carew Boulding and David S. Brown attempted to answer the question: "Does the number of political parties influence voter turnout in developing democracies?" Their research focuses on Brazil and Bolivia, concluding that differences in electoral systems shape the relationship between party competition and voter turnout. They find that in Bolivia, an increase in the number of parties is strongly associated with higher turnout, while in Brazil, the relationship is weaker and slightly negative. However, their study is limited to only two countries, which constrains the generalizability of their findings.

## Hypothesis and Theory

The Voter Turnout Model, proposed by @downs1957, has profoundly influenced theories of electoral behavior. Downs argues that voters are rational actors who weigh the costs and benefits of voting. In his model, the individual benefit of voting is effectively zero, as no single vote is likely to determine the outcome of an election. This would suggest that no one should vote, since the costs—including registration, travel to polling places, and gathering information—outweigh the expected utility. However, empirical evidence shows that substantial numbers of people do vote across democracies, indicating that additional, non-material or expressive benefits must play a role.

To better understand these dynamics, @powell1982 identifies three categories of factors that shape political participation: the socioeconomic environment, the constitutional setting, and the party system. My analysis builds on this framework by controlling for the first two in order to isolate the effect of party fragmentation, a key characteristic of party systems.

I hypothesize that the degree of party institutionalization moderates the effect of party fragmentation on voter turnout. In developing democracies with less consolidated institutions, increasing fragmentation—especially over time—can trigger skepticism among voters, as newer or smaller parties tend to be less institutionalized. The effect on turnout, however, is theoretically ambiguous and depends on voter perceptions. If voters interpret fragmentation as a sign of elite disunity or opportunism, this may erode trust in the system and reduce turnout. Alternatively, if fragmentation increases the perceived stakes of elections or enhances representation, it may stimulate greater participation. Party institutionalization refers to the extent to which parties are socially rooted, programmatically coherent, and organizationally autonomous @randall2002. In weakly institutionalized systems, high fragmentation exacerbates information costs, making it harder for voters to distinguish between poorly defined political options. As Downs (1957) suggests, when these costs outweigh perceived benefits, turnout declines. Low institutionalization may also undermine party credibility and voters' ability to make informed choices, disproportionately affecting less educated or lower-information voters.

Additionally, a proliferation of weak parties can dilute party brands and deepen citizen disillusionment, suppressing political engagement. Conversely, institutionalized parties tend to maintain clearer ideological profiles, mobilize voters through established networks, and reduce uncertainty. In such contexts, fragmentation may actually encourage participation by offering more meaningful and credible choices. That said, fragmentation can also mobilize turnout under certain conditions—for example, in highly polarized environments where voters perceive rising competition or potential threats. In these cases, even poorly institutionalized parties may benefit from heightened salience, driving voters to the polls despite high information costs.

Ultimately, my argument emphasizes that fragmentation’s effect on turnout is conditional on institutionalization. In systems with strong parties, fragmentation may reflect healthy political pluralism and enhance democratic participation. In systems with weak parties, it may instead signal volatility, confusion, or elite self-interest, leading to voter disengagement. This theoretical framework informs my empirical strategy, which includes an interaction term between fragmentation and institutionalization to test whether their combined effect influences turnout in a statistically and substantively meaningful way.

## Data and Research Design

To examine this relationship, I will use data from the Quality of Government Institute, Varieties of Democracy (V-Dem) Institute and Freedom House. From the QOG Standard Dataset 2025, the dependent variable will be measured by Voter Turnout of Parliamentary Elections (ideavt_legvt), which is available for 187 countries and the independent variable is the Effective number of electoral parties (gol_enep) according to the formula proposed by @laakso2002. This measure has been widely used in the literature and quantifies how many meaningful political parties exist in a party system by accounting for their relative sizes, including the total number of parties and their vote share. I only considered country-year observations for which both the independent and dependent variables were available. From the same dataset, I am also using a variable indicating whether the country has compulsory voting and voting age as controls.

From the Varieties of Democracy (V-Dem) General Dataset v15, I am using the variable on party institutionalization (v2xps_party). The V-Dem Party Institutionalization Index is measured from 0 to 1 and refers to various attributes of the political parties in a country, such as the level and depth of organization, links to civil society, cadres of party activists, party supporters within the electorate, coherence of party platforms and ideologies, party-line voting among representatives within the legislature. A high score on these attributes generally indicates a more institutionalized party system (V-Dem). The index is calculated by aggregating scaled indicators of other variables related to party strength, party organizations (v2psorgs), party branches (v2psprbrch), party linkages (v2psprlnks), distinct party platforms (v2psplats), and legislative party cohesion (v2pscohesv).

From the World Bank’s World Development indicators I am using variables to control for the effect of the socio-economic environment on voter turnout. In particular, I chose GDP per capita, life expectancy, total population and population density for that purpose. Literacy rates would have also been important to consider as a significant measure of human capital, but there were 85% of missing values, so the variable was removed from the analysis. Lastly, from Freedom House, I used the political rights score for each country-year observation to only run the analysis for countries with a political rights score under 3, used to classify electoral democracies.

```{r echo=FALSE}

library(dplyr)
library(modelsummary)
library(data.table)

vdem_data <- read.csv("/Users/isabellawerneck/Documents/GitHub/psci3200-camilacp03/data/V-Dem-CY-Full+Others-v15.csv")
qog_data <- read.csv("/Users/isabellawerneck/Documents/GitHub/psci3200-camilacp03/data/qog_std_ts_jan25.csv")
wdi_data <- read.csv("/Users/isabellawerneck/Documents/GitHub/psci3200-camilacp03/data/social_indicators.csv")
freedom_data <- read.csv("/Users/isabellawerneck/Documents/GitHub/psci3200-camilacp03/data/Country-Year_Freedom_Status.csv")

vdem_data_filtered <- subset(vdem_data, year >= 1980)

# Select relevant variables from QoG
qog_subset <- qog_data[, c("cname", "year", "gol_enep", "ideavt_legvt", "ideavt_legcv")]

# Rename 'cname' to match V-Dem's 'country_name' for merging
colnames(qog_subset)[colnames(qog_subset) == "cname"] <- "country_name"
colnames(freedom_data)[colnames(freedom_data) == "country"] <- "country_name"

# Left join: keep all rows from V-Dem, bring in matching QoG values
merged_data <- merge(vdem_data_filtered, qog_subset, by = c("country_name", "year"), all.x = TRUE)
merged_data <- merge(merged_data, wdi_data, by = c("country_name", "year"), all.x = TRUE)
merged_data <- merge(merged_data, freedom_data, by = c("country_name", "year"), all.x = TRUE)
merged_data <- merged_data[complete.cases(merged_data[, c("gol_enep", "ideavt_legvt")]), ]


## Renaming variables 

merged_data <- merged_data %>%
  rename(
    turnout = ideavt_legvt,
    party_strength = v2xps_party,
    party_fragmentation = gol_enep,
    compulsory_voting = ideavt_legcv)

# Convert to numeric
merged_data$population <- as.numeric(as.character(merged_data$population))
merged_data$life_expectancy <- as.numeric(as.character(merged_data$life_expectancy))
merged_data$pop_density <- as.numeric(as.character(merged_data$pop_density))
merged_data$gdp_percapita <- as.numeric(as.character(merged_data$gdp_percapita))

## Filtering by freedom score 

merged_data_F <- merged_data %>% 
  filter(freedom_status == "F")

merged_data_PF <- merged_data %>% 
  filter(freedom_status %in% c("F","PF"))

## Fixed Effects Model with Interaction Terms 

model_fe_F <- lm(turnout ~ year + factor(country_name) + party_fragmentation + party_strength + 
                 party_fragmentation * party_strength + population + life_expectancy + gdp_percapita +compulsory_voting,
               data = merged_data_F)

model_fe_PF <- lm(turnout ~ year + factor(country_name) + party_fragmentation + party_strength + 
                    party_fragmentation * party_strength + population + life_expectancy + gdp_percapita +compulsory_voting,
                  data = merged_data_PF)

# Load necessary libraries
#install.packages("tidyr") 
library(tidyr)
#install.packages("gt")       
library(dplyr)
library(gt)

# Select key variables and calculate summary statistics
summary_data <- merged_data_F %>%
  summarise(
    Turnout_Mean = mean(turnout, na.rm = TRUE),
    Turnout_SD = sd(turnout, na.rm = TRUE),
    Turnout_Min = min(turnout, na.rm = TRUE),
    Turnout_Max = max(turnout, na.rm = TRUE),
    
    Fragmentation_Mean = mean(party_fragmentation, na.rm = TRUE),
    Fragmentation_SD = sd(party_fragmentation, na.rm = TRUE),
    Fragmentation_Min = min(party_fragmentation, na.rm = TRUE),
    Fragmentation_Max = max(party_fragmentation, na.rm = TRUE),

    Strength_Mean = mean(party_strength, na.rm = TRUE),
    Strength_SD = sd(party_strength, na.rm = TRUE),
    Strength_Min = min(party_strength, na.rm = TRUE),
    Strength_Max = max(party_strength, na.rm = TRUE),

    GDP_Mean = mean(gdp_percapita, na.rm = TRUE),
    GDP_SD = sd(gdp_percapita, na.rm = TRUE),
    GDP_Min = min(gdp_percapita, na.rm = TRUE),
    GDP_Max = max(gdp_percapita, na.rm = TRUE),

    CompulsoryVoting_Mean = mean(compulsory_voting, na.rm = TRUE),
    CompulsoryVoting_SD = sd(compulsory_voting, na.rm = TRUE),
    CompulsoryVoting_Min = min(compulsory_voting, na.rm = TRUE),
    CompulsoryVoting_Max = max(compulsory_voting, na.rm = TRUE)
  ) %>%
  # Transpose to long format
  pivot_longer(everything(),
               names_to = c("Variable", "Statistic"),
               names_sep = "_") %>%
  tidyr::pivot_wider(names_from = Statistic, values_from = value)

# Display with gt
summary_data %>%
  gt() %>%
  tab_header(
    title = "Table 1: Descriptive Statistics of Key Variables"
  ) %>%
  fmt_number(columns = where(is.numeric), decimals = 2)
```

I will use a regression model with country-fixed effects and an interaction term to capture the relationship between party fragmentation and voter turnout in the context of my hypothesis. The purpose of including country-fixed effects is to account for variations in the electoral system within countries that may have influenced voter turnout over time, such as electoral reforms that significantly increased or decreased the number of parties. By holding constant cross-country variation in time-invariant unobservable factors, this methodology helps eliminate the influence of potential confounders in voter turnout, such as cultural and institutional differences between countries. Controlling for country-fixed effects thus serves as a valid proxy for the third category of voter turnout determinants identified by @powell1982: the constitutional environment. Since this environment tends to vary across countries but remains relatively stable over time within countries, fixed effects can effectively account for its impact.

In addition to the variables presented in the Data section, I will also add an interaction term to capture the combined effect of party fragmentation and party institutionalization, an important element to test the accuracy of my hypothesis. Empirically, this term allows the model to test whether the effect of party fragmentation on voter turnout depends on the level of institutionalization. That is, it examines whether a one-unit increase in fragmentation has a different impact on turnout in systems with weakly institutionalized parties versus those with highly institutionalized ones. The complete regression model can be seen below:

### Country Fixed-Effects (FE) Model

$$
\text{VoterTurnout}_{it} = \beta_1 \cdot \text{Fragmentation}_{it} + \beta_2 \cdot \text{Institutionalization}_{it} + \beta_3 \cdot \text{Institutionalization}_{it} * \text{Fragmentation}_{it} + \beta_4 \cdot \text{CompulsoryVoting}_{it} + \beta_5 \cdot \text{NGDPGrowth}_{it} + \beta_6 \cdot \text{Population}_{it} + \beta_7 \cdot \text{PopulationDensity}_{it} + \beta_8 \cdot \text{LifeExpectancy}_{it} + \alpha_i + \varepsilon_{it}
$$

where $\alpha_i$ corresponds to the country fixed effect: unobserved, time-invariant characteristics of country i.

## Results

The plot in Figure 1 shows that the relationship between party fragmentation and voter turnout is conditional on party institutionalization, based on country-year observations for countries designated as "Free" by Freedom House’s Freedom in the World Score. In systems with low party strength, the relationship is slightly upward sloping, suggesting that greater fragmentation may modestly increase turnout by offering more choices and encouraging participation in less institutionalized contexts. In medium-strength systems, the slope is relatively flat, indicating that fragmentation has little to no effect on turnout. In contrast, in highly institutionalized systems, the relationship is slightly downward sloping, implying that excessive fragmentation may reduce turnout by confusing voters or diluting party accountability. These seeming trends will be confirmed empirically through a regression model with an interaction term and country-fixed effects as empirical extensions.

```{r echo=FALSE}

library(ggplot2)
library(dplyr)

# Filter out rows with NA in any key variable
filtered_data <- merged_data %>%
  filter(
    !is.na(party_fragmentation),
    !is.na(turnout),
    !is.na(party_strength)
  ) %>%
  mutate(party_strength_group = case_when(
    party_strength < 0.33 ~ "Low",
    party_strength >= 0.33 & party_strength < 0.66 ~ "Medium",
    party_strength >= 0.66 ~ "High"
  ))

# Plot without NA values
ggplot(filtered_data, aes(x = party_fragmentation, y = turnout)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE, color = "steelblue") +
  facet_wrap(~party_strength_group, nrow = 1) +
  labs(
    title = "Figure 1: Party Fragmentation and Voter Turnout, Conditioned on Party Institutionalization",
    x = "Effective Number of Parties (Fragmentation)",
    y = "Voter Turnout (%)"
  ) +
  theme_minimal()

```

The results of the fixed-effects regression model presented in Table 2 provide insight into how party system characteristics influence voter turnout across countries. The coefficient on party fragmentation is positive but statistically insignificant, suggesting that, on average, a higher number of electoral parties does not have a consistent or robust effect on turnout when other factors are held constant. Similarly, institutionalization (party_strength) also exhibits an insignificant main effect, indicating that its direct influence on turnout may not be uniform across contexts. However, the theoretical contribution of this model lies in the interaction between fragmentation and institutionalization, which yields a negative coefficient and approaches marginal statistical significance, though it cannot be considered a meaningful relationship once it is bellow that thereshold. The fact that the coefficients are not statistically significant is likely not related to statistical power, since the sample is large, including 571 elections in electoral democracies from 1980 to 2024.

Several control variables perform as expected. Compulsory voting is strongly and positively associated with turnout, consistent with prior findings in the literature. GDP per capita also has a significant positive effect, suggesting that economic development supports electoral participation. The year variable is negative and significant, indicating a general downward time trend in turnout, perhaps reflecting growing global disillusion with political institutions. Population size is also significant and positive, though its substantive impact is small. Other controls, such as life expectancy, do not show statistically significant effects in this model.

```{r echo=FALSE}

modelsummary(model_fe_F,
             statistic = "std.error",
             stars = TRUE,
             coef_omit = "country_name",
             output = "html",
             title = "Table 2: Fixed Effects Regression on Voter Turnout")
```

This analysis was conducted in 571 country-year observations only of countries with a designation of 'Free' on Freedom House's Freedom in the World Score. However, when running the same regression also including countries designated as 'Partially Free', the results differ. This fixed-effects model, shown in Table 3, provides empirical support for the hypothesis that the relationship between party fragmentation and voter turnout is moderated by the level of party institutionalization. The coefficient for party fragmentation is positive and statistically significant, indicating that, on average, increased fragmentation is associated with higher voter turnout. However, this effect is conditional: the interaction term between party fragmentation and party strength is negative and significant, suggesting that the positive effect of fragmentation on turnout diminishes as party institutionalization increases. In weakly institutionalized systems, fragmentation may signal political volatility or heighten electoral competition, mobilizing voters. In contrast, in systems with stronger parties, additional fragmentation may create confusion or dilute party identity, thus dampening turnout. When including countries that are considered partially free and which are oftentimes hybrid systems or transitioning to democracy, the theorized relationship between voter turnout and fragmentation can be observed empirically.

```{r echo=FALSE}

modelsummary(model_fe_F,
             statistic = "std.error",
             stars = TRUE,
             coef_omit = "country_name",
             output = "html",
             title = "Table 3: Fixed Effects Regression on Voter Turnout including Partially Free Countries")

```

## Discussion and Policy Implications

Despite the insights provided by the fixed-effects regression model, some limitations constrain the strength of causal inference in this analysis. Most notably, the model remains observational and non-experimental, making it vulnerable to unobserved time-varying confounders that may jointly influence both party system characteristics and voter turnout. While the use of country-fixed effects controls for all time-invariant country-specific factors—such as political culture or historical legacies, it does not address potential confounders that change over time, such as shifts in electoral laws, which may bias estimates. Furthermore, reverse causality cannot be ruled out. For instance, lower turnout may itself incentivize party elites to restructure the party system or weaken party institutionalization. Finally, measurement error in the institutionalization index could attenuate estimated effects. The index is constructed by V-Dem based on surveys with the local population, and largely relies on individual perceptions. As such, while the results offer theoretically consistent and suggestive patterns, they should be interpreted as associational, rather than definitively causal.

Future research could build on this analysis by adopting quasi-experimental designs to better identify causal effects. One promising approach would be to use a difference-in-differences framework, comparing countries that have undergone electoral reforms resulting in increased party fragmentation with similar countries that did not experience such changes for the control group. For example, a country that shifted from a majoritarian to a proportional electoral system, increasing the effective number of parties, could be compared to a neighboring or institutionally similar country where the system remained unchanged. Another strong approach would be to employ an instrumental variables (IVs) model. The empirical extension of instrumental variables tries to minimize endogeneity further such as by addressing reverse causality and simultaneity bias. A concern that was addressed by @valentim2024 is the endogeneity of political parties that are specific to the political context of the countries in which they operate. The authors argue that this endogeneity may bias results and contribute to the ambiguity of results in empirical scholarly research on party fragmentation and voter turnout. Therefore, I am also considering a regression model that includes the use of an Instrumental Variable to mitigate the described endogeneity risk. This could be a variable that affects party fragmentation but not directly related to voter turnout.

The findings from this study carry important implications for policy. In regimes where party institutionalization is weak, having a higher number of effective parties can increase voter turnout. Civil society groups and politicians invested in increasing electoral engagement should therefore consider advocating for greater party diversification. In more established systems with higher levels of institutionalization, the effect of fragmentation on turnout is reduced. Since this result was only statistically significant when partially free countries were included, it may be that fragmentation encourages turnout mainly in hybrid or transitioning regimes, where weak institutions make new parties more salient and hopes for change more compelling. In more stable systems with entrenched parties, the perceived value of additional parties may be lower. When considering the regression results excluding partially free countries, the effects of party fragmentation and institutionalization are not significant, suggesting a more limited relationship in fully democratic settings.

The strong correlation between compulsory voting and turnout highlights a clear opportunity to increase electoral participation through institutional reform. Additionally, the statistical significance of several country coefficients under fixed effects indicates that there are meaningful country-specific factors associated with higher or lower turnout. More research could be done to understand what regional or contextual characteristics contribute to these patterns and how they can inform policies to strengthen democratic engagement.

## References
